---
title: "NYPD Shooting Incident Data Historic Final Project"
author: "Benjamin Janes"
date: "`r Sys.Date()`"
output:  html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(flextable, lubridate, gt, tidyverse, ggplot2, ggmap, leaflet, leaflet.extras, sf, osmdata, randomForest, forecast, Rtsne)
```

1.  Start an Rmd document that describes and imports the shooting project data set in a reproducible manner.

```{r, Import_Data}

nypd_shooting_data <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

2.  Add to your Rmd document a summary of the data and clean up your data set by changing appropriate variables to factor and date types and getting rid of any columns not needed. Show the summary of your data to be sure there is no missing data. If there is missing data, describe how you plan to handle it.

<!-- -->

a.  Missing data strategy.
    For categorical variables, I will replace blank, missing, or null values with unknown or unstated.
    I will exclude missing latitude and longitude values during mapping.

b.  Imported data summary

```{r}
glimpse(nypd_shooting_data)
summary(nypd_shooting_data)
number_rows <- nrow(nypd_shooting_data)

percent_loc_of_occur <- sum(nypd_shooting_data$LOC_OF_OCCUR_DESC == "" | nypd_shooting_data$LOC_OF_OCCUR_DESC == "(null)" | is.na(nypd_shooting_data$LOC_OF_OCCUR_DESC)) / number_rows

percent_of_loc_desc <- sum(nypd_shooting_data$LOCATION_DESC == "" | nypd_shooting_data$LOCATION_DESC == "(null)" | is.na(nypd_shooting_data$LOCATION_DESC)) / number_rows

percent_of_loc_class <- sum(nypd_shooting_data$LOC_CLASSFCTN_DESC == "" | nypd_shooting_data$LOC_CLASSFCTN_DESC == "(null)" | is.na(nypd_shooting_data$LOC_CLASSFCTN_DESC)) / number_rows

percent_of_perp_race <- sum(nypd_shooting_data$PERP_RACE == "" | nypd_shooting_data$PERP_RACE == "(null)" | is.na(nypd_shooting_data$PERP_RACE)) / number_rows

percent_of_vic_race <- sum(nypd_shooting_data$VIC_RACE == "" | nypd_shooting_data$VIC_RACE == "(null)" | is.na(nypd_shooting_data$VIC_RACE)) / number_rows

percent_of_vic_race <- sum(nypd_shooting_data$VIC_RACE == "" | nypd_shooting_data$VIC_RACE == "(null)" | is.na(nypd_shooting_data$VIC_RACE)) / number_rows

percent_of_vic_age_group <- sum(nypd_shooting_data$VIC_AGE_GROUP == "" | nypd_shooting_data$VIC_AGE_GROUP == "(null)" | is.na(nypd_shooting_data$VIC_AGE_GROUP)) / number_rows

percent_of_perp_age_group <- sum(nypd_shooting_data$PERP_AGE_GROUP == "" | nypd_shooting_data$PERP_AGE_GROUP == "(null)" | is.na(nypd_shooting_data$PERP_AGE_GROUP)) / number_rows


```

b.  Define level values converting variables from a character to factor type.

```{r}
# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" 

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the BORO variable.

boro_options <<- nypd_shooting_data %>%
  mutate(BORO = ifelse(is.na(BORO) | BORO == "" | BORO == "(null)", "UNKNOWN", BORO)) %>%
  select(BORO) %>%
  distinct() %>%
  pull()  # formats variable as a vector that can be read by the factor transformation
# boro_options

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the JURISDICTION_CODE variable.

jurisdiction_code_options <- nypd_shooting_data %>%
  mutate(JURISDICTION_CODE = ifelse(is.na(JURISDICTION_CODE) | JURISDICTION_CODE == "" | JURISDICTION_CODE == "(null)", "UNKNOWN", JURISDICTION_CODE)) %>%
  select(JURISDICTION_CODE) %>%
  distinct() %>%
  pull(JURISDICTION_CODE) # formats variable as a vector that can be read by the factor transformation

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the LOC_OF_OCCUR_DESC variable.

loc_of_occur_desc_options <- nypd_shooting_data %>%
  mutate(LOC_OF_OCCUR_DESC = ifelse(is.na(LOC_OF_OCCUR_DESC) | LOC_OF_OCCUR_DESC == "" | LOC_OF_OCCUR_DESC == "(null)", "UNKNOWN", LOC_OF_OCCUR_DESC)) %>%
  select(LOC_OF_OCCUR_DESC) %>%
  distinct() %>%
  pull(LOC_OF_OCCUR_DESC) # formats variable as a vector that can be read by the factor transformation

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the PERP_RACE variable.

perp_race_options <- nypd_shooting_data %>%
  mutate(PERP_RACE = ifelse(is.na(PERP_RACE) | PERP_RACE == "" | PERP_RACE == "(null)", "UNKNOWN", PERP_RACE)) %>%
  select(PERP_RACE) %>%
  distinct() %>%
  pull(PERP_RACE)

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the VIC_RACE variable.

vic_race_options <- nypd_shooting_data %>%
  mutate(VIC_RACE = ifelse(is.na(VIC_RACE) | VIC_RACE == "", "UNKNOWN", VIC_RACE)) %>%
  select(VIC_RACE) %>%
  distinct() %>%
  pull(VIC_RACE)

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the VIC_AGE_GROUP variable.

vic_age_options <- nypd_shooting_data %>%
  mutate(VIC_AGE_GROUP = ifelse(is.na(VIC_AGE_GROUP) | VIC_AGE_GROUP == "", "UNKNOWN", VIC_AGE_GROUP)) %>%
  select(VIC_AGE_GROUP) %>%
  distinct() %>%
  pull(VIC_AGE_GROUP)

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the PERP_AGE_GROUP variable.

perp_age_options <- nypd_shooting_data %>%
  mutate(PERP_AGE_GROUP = ifelse(is.na(PERP_AGE_GROUP) | PERP_AGE_GROUP == "", "UNKNOWN", PERP_AGE_GROUP)) %>%
  select(PERP_AGE_GROUP) %>%
  distinct() %>%
  pull(PERP_AGE_GROUP)

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the LOC_CLASSFCCTN_DESC variable.

loc_class_desc_options <- nypd_shooting_data %>%
  select(LOC_CLASSFCTN_DESC) %>%
  mutate(LOC_CLASSFCTN_DESC = ifelse(is.na(LOC_CLASSFCTN_DESC) | LOC_CLASSFCTN_DESC == "" | LOC_CLASSFCTN_DESC == "(null)", "UNKNOWN", LOC_CLASSFCTN_DESC)) %>%
  distinct() %>%
  pull()
# loc_class_desc_options

# Create a distinct list of values for each factor variable and label all missing data as, "UNKNOWN" for the LOCATION_DESC variable.

loc_desc_options <- nypd_shooting_data %>%
  select(LOCATION_DESC) %>%
  mutate(LOCATION_DESC = ifelse(is.na(LOCATION_DESC) | LOCATION_DESC == "" | LOCATION_DESC == "(null)", "UNKNOWN", LOCATION_DESC)) %>%
  distinct() %>%
  pull()
# loc_desc_options
```

c.  Apply transformations to main data as a new data set nypd_shooting_data_tidy.\

```{R}

nypd_shooting_data_tidy <- nypd_shooting_data %>%
  # Eliminate variables that will not be used
  select(-c(INCIDENT_KEY, X_COORD_CD, Y_COORD_CD, Lon_Lat, STATISTICAL_MURDER_FLAG)) %>%

  # Convert date and time values
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    OCCUR_TIME = hms(OCCUR_TIME)
  ) %>%

  # Convert BORO to a factor variable
  mutate(
    BORO = ifelse(is.na(BORO) | BORO == "" | BORO == "(null)", "UNKNOWN", BORO),
    BORO = factor(BORO, levels = boro_options)
  ) %>%

  # Convert JURISDICTION_CODE to a factor variable
  mutate(
    JURISDICTION_CODE = ifelse(is.na(JURISDICTION_CODE) | JURISDICTION_CODE == "" | JURISDICTION_CODE == "(null)", "UNKNOWN", JURISDICTION_CODE),
    JURISDICTION_CODE = factor(JURISDICTION_CODE, levels = jurisdiction_code_options)
  ) %>%

  # Convert LOC_OF_OCCUR_DESC to a factor variable
  mutate(
    LOC_OF_OCCUR_DESC = ifelse(is.na(LOC_OF_OCCUR_DESC) | LOC_OF_OCCUR_DESC == "" | LOC_OF_OCCUR_DESC == "(null)", "UNKNOWN", LOC_OF_OCCUR_DESC),
    LOC_OF_OCCUR_DESC = factor(LOC_OF_OCCUR_DESC, levels = loc_of_occur_desc_options)
  ) %>%

  # Convert sex to a factor variable
  mutate(
    PERP_SEX = factor(
      recode(toupper(trimws(PERP_SEX)),
             "M" = "M", "F" = "F", .default = "Not Stated"),
      levels = c("M", "F", "Not Stated")
    ),
    VIC_SEX = factor(
      recode(toupper(trimws(VIC_SEX)),
             "M" = "M", "F" = "F", .default = "Not Stated"),
      levels = c("M", "F", "Not Stated")
    )
  ) %>%

  # Convert race to a factor variable
  mutate(
    VIC_RACE = ifelse(is.na(VIC_RACE) | VIC_RACE == "" | VIC_RACE == "(null)", "UNKNOWN", VIC_RACE), 
    VIC_RACE = factor(VIC_RACE, levels = vic_race_options),
    PERP_RACE = ifelse(is.na(PERP_RACE) | PERP_RACE == "" | PERP_RACE == "(null)", "UNKNOWN", PERP_RACE), 
    PERP_RACE = factor(PERP_RACE, levels = perp_race_options)
  ) %>%

  # Convert age groups to a factor variable
  mutate(
    VIC_AGE_GROUP = ifelse(is.na(VIC_AGE_GROUP) | VIC_AGE_GROUP == "" | VIC_AGE_GROUP == "(null)", "UNKNOWN", VIC_AGE_GROUP), 
    VIC_AGE_GROUP = factor(VIC_AGE_GROUP, levels = vic_age_options),
    PERP_AGE_GROUP = ifelse(is.na(PERP_AGE_GROUP) | PERP_AGE_GROUP == "" | PERP_AGE_GROUP == "(null)", "UNKNOWN", PERP_AGE_GROUP), 
    PERP_AGE_GROUP = factor(PERP_AGE_GROUP, levels = perp_age_options)
  ) %>%

  # Convert location classifications to a factor variable
  mutate(
    LOC_CLASSFCTN_DESC = ifelse(
      is.na(LOC_CLASSFCTN_DESC) | LOC_CLASSFCTN_DESC == "" | LOC_CLASSFCTN_DESC == "(null)", 
      "UNKNOWN", 
      LOC_CLASSFCTN_DESC
    ),
    LOC_CLASSFCTN_DESC = factor(LOC_CLASSFCTN_DESC, levels = loc_class_desc_options),
    
    LOCATION_DESC = ifelse(
      is.na(LOCATION_DESC) | LOCATION_DESC == "" | LOCATION_DESC == "(null)", 
      "UNKNOWN", 
      LOCATION_DESC
    ),
    LOCATION_DESC = factor(LOCATION_DESC, levels = loc_desc_options)
  )

```

d.  Summary of tidy data.

```{r}
glimpse(nypd_shooting_data_tidy)
summary(nypd_shooting_data_tidy)
```

3.  Conduct a clustering analysis to idenitfy common factors across shooting incidents.

The clustering analysis grouped shooting incidents into four distinct categories based on shared characteristics like location type, borough, and victim-perpetrator demographics.
These groupings highlight distinct patterns in shooting incidents, which can help focus resources and interventions based on location and context.

The first cluster represents incidents that predominantly occurred in high-density urban areas, often outdoors.
Potential interventions include

1.  Inreased police patrols in high-density urban areas, particularly during peak hours for outdoor activity.

2.  Improved community engagement by partnering with local organizations to address underlying social issues, such as gang activity or youth violence, in these areas.

3.  Improve environmental designs such as increased urban lighting and removing abandoned structures.

The second cluster is associated with residential neighborhoods, showing a mix of victim and perpetrator demographics.
Potential interventions include

1.  Implementing conflict resolution and mediation programs to defuse domestic disputes or community conflicts in residential areas.

2.  Increase neighborhood watch programs that empower residents to take collective action for safety with police support.

3.  Increase social services access to counseling, housing assistance, or substance abuse treatment services and address the root causes of violence.

The third cluster primarily consists of indoor incidents involving male victims.Potential Interventions:

1.  Improve targeted crime prevention by workign with property managers and landlords to install surveillance cameras and improve security in indoor settings.

2.  Increase youth outreach mentoring prograns and assistance in finding employment opportunities for young men.

3.  Field crisis response teams that can provide rapid mental health or conflict resolution support to address interpersonal violence in indoor settings.

The fourth cluster is smaller and contains unique or sparse incidents that don’t align with the patterns of the other clusters.
Potential interventions incldue

1.  Improved hotspot analysis by conducting detailed investigations into incidents to identify unique causes (e.g., targeted violence, organized crime).

2.  Increased data-driven policing leveraging predictive analytics to deploy resources dynamically in areas with sporadic but severe incidents.

3.  Conductiing tailored interventions by collaborating with local influencers to develop solutions tailored to the specific areas or communities where shootings occur.

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(cluster)
library(Rtsne)

# Assuming `nypd_shooting_data_tidy` is your cleaned dataset
# Select contextual variables for clustering
clustering_data <- nypd_shooting_data_tidy %>%
  select(BORO, LOC_OF_OCCUR_DESC, JURISDICTION_CODE, VIC_SEX, PERP_SEX) %>%
  mutate_all(as.factor)

# Convert categorical variables into numeric format using one-hot encoding
clustering_data_encoded <- model.matrix(~ . - 1, data = clustering_data) %>%
  as.data.frame()

# Remove duplicate rows for clustering
clustering_data_unique <- clustering_data_encoded[!duplicated(clustering_data_encoded), ]

# Scale the data to normalize values
clustering_data_scaled <- scale(clustering_data_unique)

# Compute a distance matrix
distance_matrix <- dist(clustering_data_scaled, method = "euclidean")

# Perform hierarchical clustering using Ward's method
hc <- hclust(distance_matrix, method = "ward.D2")

# Cut the dendrogram into clusters (e.g., 4 clusters)
clusters <- cutree(hc, k = 4)

# Add cluster labels to unique rows
clustering_data_unique$Cluster <- as.factor(clusters)

# Map cluster assignments back to the original dataset
clustering_data_encoded$Cluster <- clustering_data_unique$Cluster[match(
  apply(clustering_data_encoded, 1, paste, collapse = ""),
  apply(clustering_data_unique[, -ncol(clustering_data_unique)], 1, paste, collapse = "")
)]

# Add cluster labels to the original dataset
nypd_shooting_data_tidy$Cluster <- clustering_data_encoded$Cluster

# Summarize characteristics of each cluster
cluster_summary <- nypd_shooting_data_tidy %>%
  group_by(Cluster) %>%
  summarize(
    Count = n(),
    Common_Borough = names(sort(table(BORO), decreasing = TRUE)[1]),
    Common_Location = names(sort(table(LOC_OF_OCCUR_DESC), decreasing = TRUE)[1]),
    Common_Jurisdiction = names(sort(table(JURISDICTION_CODE), decreasing = TRUE)[1]),
    Common_Victim_Sex = names(sort(table(VIC_SEX), decreasing = TRUE)[1]),
    Common_Perpet_Sex = names(sort(table(PERP_SEX), decreasing = TRUE)[1])
  )

# Print cluster summary
print(cluster_summary)

# Perform t-SNE for dimensionality reduction
tsne_results <- Rtsne(clustering_data_scaled, perplexity = 30, theta = 0.0)

# Convert t-SNE results to a data frame
tsne_data <- data.frame(tsne_results$Y, Cluster = clustering_data_unique$Cluster)

# Define cluster labels for interpretation
cluster_labels <- c(
  "Cluster 1: High urban density, outdoor incidents",
  "Cluster 2: Residential areas, mixed demographics",
  "Cluster 3: Predominantly indoors, male victims",
  "Cluster 4: Sparse data, unique outliers"
)
```

```{r fig.width=10, fig.height=12}
# Plot clusters
ggplot(tsne_data, aes(x = X1, y = X2, color = Cluster)) +
  geom_point(alpha = 0.8, size = 2) +  # Slightly smaller points for better visibility
  labs(
    title = "Hierarchical Clustering of Shooting Incidents (t-SNE Dimensions)",
    x = "Dimension 1\n(Shared Characteristics: Locations, Demographics)",
    y = "Dimension 2\n(Differentiating Factors: Perpetrator and Victim Attributes)",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12) +  # Adjust base font size for readability
  scale_color_manual(
    values = c("red", "blue", "green", "purple"),
    labels = c(
      "Cluster 1: High urban density, outdoor incidents",
      "Cluster 2: Residential areas, mixed demographics",
      "Cluster 3: Predominantly indoors, male victims",
      "Cluster 4: Sparse data, unique outliers"
    )
  ) +
  theme(
    legend.position = "bottom",  # Move legend below the plot
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 9),
    legend.box = "vertical",  # Arrange legend items vertically
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),  # Center the title
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 9),
    plot.margin = unit(c(1, 1, 3, 1), "cm")  # Add extra bottom margin for the vertical legend
  ) +
  guides(
    color = guide_legend(ncol = 1)  # Force legend into 1 column (vertical layout)
  )
```

4.  Add at least two different visualizations & some analysis to your Rmd. Does this raise additional questions that you should investigate?

<!-- -->

a.  Additional questions:

-   Why is the race unknown for some of these shootings?
-   Where are the police officers concentrated?
-   Could we place police officers in areas where shooting occur at a higher rate than other areas?
-   Other visuals of the data mapping incidents by borough, time of day, and location type
-   Relating the shooting incident to time of year, seasons, and holidays to determine if these are predictors of shooting incidents

b.  This is a heat map of overlay of the New York City showing where the shootings occur.

```{r out.width='100%', out.height='700px', eval = knitr::is_html_output()}

nypd_shooting_data_map <- nypd_shooting_data_tidy %>%
  filter(!is.na(Latitude), !is.na(Longitude))

leaflet(data = nypd_shooting_data_map) %>%
  addTiles() %>%  # Default OpenStreetMap layer
  setView(lng = -73.91, lat = 40.74, zoom = 11) %>%  # Center on NYC
  addHeatmap(
    lng = ~Longitude, lat = ~Latitude, 
    intensity = ~1,  # Use 1 for uniform heat intensity
    blur = 20, max = 0.05, radius = 15  # Adjust heat map appearance
  ) %>%
  addLegend("bottomright", colors = c("blue", "red"), labels = c("Low", "High"),
            title = "Shooting Incident Density") %>%
  addProviderTiles(providers$OpenStreetMap)

```

c.  Race comparison of the victim and perpetrator.

```{r out.width='100%', out.height='700px'}

ggplot(nypd_shooting_data_tidy) +
  geom_bar(aes(x = VIC_RACE, fill = "Victim"), alpha = 0.6, position = "dodge") +
  geom_bar(aes(x = PERP_RACE, fill = "Perpetrator"), alpha = 0.6, position = "dodge") +
  scale_fill_manual(values = c("Victim" = "purple", "Perpetrator" = "brown")) +
  labs(title = "Race Distribution of Victims and Perpetrators", x = "Race", y = "Count") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

5.  Possible sources of bias.

-   Reporting bias. The data may not capture all shooting incidents due to under reporting or data collection inconsistencies. Data collection overall was not very good as there were significant gaps in the data that were filled in with “UNKNOWN.”
-   Selection bias. A higher number of shooting incidents may be reported where there are concentration of police officers. Adding police officer deployments would provide additional insights into this potential bias.
-   Time bias. The collected data represents a snapshot in time which may or may not be consistent with long-term results. Continuous analysis of both longer and shorter time frames would be required to compensate for this bias.
-   Data collection methods. Due the large amount of missing data, there may be significant differences in how and what shooting incident data is captured as well as the priority to collect the data as part of the investigation process across precincts and boroughs.

5.  Conclusion. The clustering analysis and heatmap visualization provide a comprehensive overview of shooting incidents in the city, emphasizing distinct patterns in location, context, and demographics. The heatmap highlights areas with high concentrations of shootings, particularly in densely populated urban zones, aligning with Cluster 1, where interventions such as increased police patrols, community engagement, and environmental design improvements are critical. Residential neighborhoods (Cluster 2) and indoor settings with male victims (Cluster 3) require targeted approaches like neighborhood watch programs, counseling services, and improved property security. Sparse, unique incidents in Cluster 4 underline the importance of hotspot analysis and data-driven policing. However, the analysis also raises important questions about the impact of reporting biases, officer deployments, and missing data on the findings. These insights underscore the need for tailored, data-informed interventions and further exploration of spatial, temporal, and contextual factors influencing shooting incidents.
